Experiences with Planning for Natural Language Generation

### 2009-06-03 trunk/Papers/ci-spark/domains.tex
changes to section 4

Figure~\ref{fig:white-rabbit-as-planning} shows the
transformed planning operators needed to generate the above example
sentence, ``The white rabbit sleeps.'' Here the action instance
$\addact{sleeps}(\mathsf{root}, e, r_1)$ replaces the atom
$\mathsf{subst}(S,\mathsf{root})$ with the atom
$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$. In an abuse of PDDL
syntax, we write $\mathsf{subj}(\mathsf{root})$ as a shorthand for a fresh
individual name.\footnote{These terms are not valid in
  ordinary PDDL but can be eliminated by estimating an upper bound $n$
  for the plan length, making $n$ copies of each action, ensuring that
  copy $i$ can only be applied in step $i$, and replacing the term
  $\mathsf{subj}(u)$ in an action copy by the constant
  $\mathsf{subj}_i$. The terms $S$, $NP$, and $N$ in the planning
  problem are constants.}  At the same time, the operator records that
the semantic information $\mathsf{sleep}(e,r_1)$ has now been
expressed, and introduces all individuals except $r_1$ as
distractors for the new RE at $\mathsf{subj}(\mathsf{root})$. These
distractors can then be removed by subsequent applications of the
other two operators. Eventually we reach a goal state, which is
characterized by goals including $\forall x \forall y. \neg
\mathsf{subst}(x,y)$, $\forall x \forall y. \neg
\mathsf{distractor}(x,y)$, and
$\mathsf{expressed}(\mathsf{sleep},e,r_1)$. For instance, the following
plan correctly performs the necessary derivation:
%
\begin{enumerate}
\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
\todo{FIX: add-sleeps takes 3 arguments in the PDDL spec!}
\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
\end{enumerate}
%
The grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps,'' can be systematically
reconstructed from this plan. Thus, we can solve the sentence
generation problem via the detour through planning and bring current
search heuristics for planning to bear on generation.

>>>>

Figure~\ref{fig:white-rabbit-as-planning} shows the
transformed planning operators needed to generate the above example
sentence, ``The white rabbit sleeps.'' Here the action instance
$\addact{sleeps}(\mathsf{root}, e, r_1)$ replaces the atom
$\mathsf{subst}(S,\mathsf{root})$ with the atom
$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$. In an abuse of PDDL
syntax, we write $\mathsf{subj}(\mathsf{root})$ as a shorthand for a fresh
individual name.\footnote{These terms are not valid in
  ordinary PDDL but can be eliminated by estimating an upper bound $n$
  for the plan length, making $n$ copies of each action, ensuring that
  copy $i$ can only be applied in step $i$, and replacing the term
  $\mathsf{subj}(u)$ in an action copy by the constant
  $\mathsf{subj}_i$. The terms $S$, $NP$, and $N$ are constants.}
At the same time, the operator records that
the semantic information $\mathsf{sleep}(e,r_1)$ has now been
expressed, and introduces all individuals except $r_1$ as
distractors for the new RE at $\mathsf{subj}(\mathsf{root})$. These
distractors can then be removed by subsequent applications of the
other two operators. Eventually we reach a goal state, which is
characterized by goals including $\forall x \forall y. \neg
\mathsf{subst}(x,y)$, $\forall x \forall y. \neg
\mathsf{distractor}(x,y)$, and
$\mathsf{expressed}(\mathsf{sleep},e,r_1)$. For instance, the following
plan correctly performs the necessary derivation:
%
\begin{enumerate}
\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
\todo{FIX: add-sleeps takes 3 arguments in the PDDL spec!}
\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
\end{enumerate}
%
The grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps,'' can be systematically
reconstructed from this plan. Thus, we can solve the sentence
generation problem via the detour through planning and bring current
search heuristics for planning to bear on generation.

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
changes to section 4

We will now return to the original question of the paper: Is planning
technology ready for applications in natural language generation? We
will tackle this question by evaluating the performance of several
planners on the NLG planning domains from the previous section.

We start with two experiments in the CRISP domain. First, we present a
scenario which focuses on the generation of referring expressions with
a tiny grammar (Section~\ref{sec:exper-1:-sent}); then we look at a
setting in which CRISP is used for surface realization with the XTAG
Grammar \citep{xtag01:_tr}, a large-scale TAG grammar of English
(Section~\ref{sec:exper-2:-sent-xtag}).

We then go through a series of experiments in the GIVE domain. We
start with a domain that is essentially Gridworld
(Section~\ref{sec:exper-2:-minim}), and then add positions that are
either not needed (Section~\ref{sec:experiment-3:-give}) or even
inaccessible (Section~\ref{sec:experiment-4:-give}). This series of
experiments is set up to illustrate what we perceive as the main
limitation of current off-the-shelf planners for our applications:
They spend a long time on computing all ground instances, even if most
of these instances are never used in the plan search. \todo{Ron, this
  is something you need to adapt to the new story -- AK}

@@

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}.\footnote{All runtimes in
  Sections~\ref{sec:exper-1:-sent} and \ref{sec:exper-2:-sent-xtag}
  were measured on a single core of an AMD Opteron 8220 CPU running at
  2.8 GHz, under Linux. FF 2.3 and Metric-FF were recompiled as 64-bit
  binaries and run with a memory limit of 32 GB. Java programs were
  executed under Java 1.6.0\_13 in 64-bit mode. They were allowed to
  ``warm up'', i.e. the JVM was given the opportunity to just-in-time
  compile the relevant bytecode by running the planner three times and
  discarding the runtimes before taking the actual measurements. All
  runtimes are averaged over three runs of the planners.} The input
parameters $(m,n)$ are plotted in lexicographic order on the
horizontal axis, and the runtime is shown in seconds on the vertical
axis, on a logarithmic scale. These results reveal a number of
interesting insights. First, the search times of FF and Metric-FF
(shown as thinner lines) significantly outperform SGPLAN's search in
this domain---on the largest instances, by a factor of over
100.\footnote{For FF and Metric-FF, we report the ``searching'' and
  ``total'' times reported by the planners. For SGPLAN, we report the
  ``total'' time and the difference between the ``total'' and
  ``parsing'' times.} Second, FF and Metric-FF perform very similarly
to each other, and their search times are almost the same as those of
the SPUD algorithm, which is impressive because they are complete
search algorithms, whereas SPUD's greedy algorithm is not.

@@

\begin{figure}[t]
\subfloat[$k=1$]{\includegraphics[width=8cm]{data/xtag-k1.pdf}}
\subfloat[$k=2$]{\includegraphics[width=8cm]{data/xtag-k2.pdf}}
  \caption{Results for the XTAG experiment, at $k=1$ and $k=2$.}
  \label{fig:xtag-graph}
\end{figure}

>>>>

We now return to the original question of the paper: is planning technology
ready for realistic applications in natural language generation? To investigate
this question we consider two sets of experiments, designed to evaluate the
performance of several planners on the NLG planning domains from the previous
section. Starting with the CRISP domain, we first present a scenario which
focuses on the generation of referring expressions with a tiny grammar
(Section~\ref{sec:exper-1:-sent}). We then look at a setting in which CRISP is
used for surface realization with the XTAG Grammar \citep{xtag01:_tr}, a
large-scale TAG grammar for English (Section~\ref{sec:exper-2:-sent-xtag}). In
the second set of experiments we investigate the GIVE domain. We begin with a
domain that is similar to the classic Gridworld
(Section~\ref{sec:exper-2:-minim}), and then add extra grid cells to the world
that are neither needed (Section~\ref{sec:experiment-3:-give}) nor even
accessible (Section~\ref{sec:experiment-4:-give}). We also investigate the role
that goal ordering plays in some of these problems. These experiments are
configured in a way that lets us explore the scalability of a planner's search
and preprocessing capabilities, and illustrates what we perceive to be one of
the main limitations of current off-the-shelf planners for our applications:
they often spend a long time computing ground instances, even when most of
these instances are not required during plan search.

@@

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}.\footnote{All runtimes in
  Sections~\ref{sec:exper-1:-sent} and \ref{sec:exper-2:-sent-xtag}
  were measured on a single core of an AMD Opteron 8220 CPU running at
  2.8 GHz, under Linux. FF 2.3 and Metric-FF were recompiled as 64-bit
  binaries and run with a memory limit of 32 GB. Java programs were
  executed under Java 1.6.0\_13 in 64-bit mode and were allowed to
  ``warm up'', i.e., the JVM was given the opportunity to just-in-time
  compile the relevant bytecode by running the planner three times and
  discarding the runtimes before taking the actual measurements. All
  runtimes are averaged over three runs of the planners.} The input
parameters $(m,n)$ are plotted in lexicographic order on the
horizontal axis, and the runtime is shown in seconds on the vertical
axis, on a logarithmic scale. These results reveal a number of
interesting insights. First, the search times of FF and Metric-FF
(shown as thinner lines) significantly outperform SGPLAN's search in
this domain---on the largest instances, by a factor of over
100.\footnote{For FF and Metric-FF, we report the ``searching'' and
  ``total'' times reported by the planners. For SGPLAN, we report the
  ``total'' time and the difference between the ``total'' and
  ``parsing'' times.} Second, FF and Metric-FF perform very similarly
to each other, and their search times are almost the same as those of
the SPUD algorithm, which is impressive because they are complete
search algorithms, whereas SPUD's greedy algorithm is not.

@@

\begin{figure}[t]
\subfloat[$k=1$]{\includegraphics[width=7cm]{data/xtag-k1.pdf}}
\subfloat[$k=2$]{\includegraphics[width=7cm]{data/xtag-k2.pdf}}
  \caption{Results for the XTAG experiment, at $k=1$ and $k=2$.}
  \label{fig:xtag-graph}
\end{figure}
